---
title: "Lake Eutrophication and Water Quality Classification Model"
author: "Lauren Chetley & Emma Rodriguez"
date: today
format: 
  html:
    theme: flatly
    toc: true
    toc-title: "🌱 Contents"
    toc-depth: 3
    number-sections: true
    code-fold: true
    code-summary: "Show Code"
    fig-align: center
    fig-width: 9
    fig-height: 6
    backgroundcolor: "#e8f5e1"
    fontcolor: "#1b4332"
    linkcolor: "#40916c"
    monobackgroundcolor: "#d8f3dc"
    fontsize: 1.1em
    linestretch: 1.8
    mainfont: "Georgia"
editor: visual
resources:
  - figures/flowchart.jpg
  - figures/flowchart.pdf
---

## Lake Water Quality Contract

### Modules Parameters

#### Nutrients (Phosphorus and Nitrogen) Module

Input: Daily lake phosphorus and nitrogen nutrient inputs (mg/L)

-   Processing: Daily values aggregated to monthly means by water year (October - September)

-   Function: nutrient_function

Output: Total Lake Productivity Risk Classifications:

Phosphorus = P and Nitrogen = N

-   Oligotrophic: Low Productivity: P \< 0.01 mg/L, N \< 0.35 mg/L

-   Mesotrophic: Medium Productivity: P 0.01 - 0.03 mg/L, N 0.36-0.65 mg/L

-   Eutrophic: High Productivity: P 0.03 - 0.10 mg/L, N 0.65 - 1.20 mg/L

-   Hypereutrophic: Very High Productivity: P \> 0.1 mg/L, N \> 1.20 mg/L

#### 

#### Dissolved Oxygen Module

Input: Daily lake dissolved oxygen levels (mg/L)

-   Processing: Daily values aggregated to monthly means by water year (October - September)

-   Function: dissoxygen_function

Output: Mean Lake Dissolved Oxygen (DO) Classifications:

-   Excellent DO: \> 8 mg/L

-   Moderate DO: 5 - 8 mg/L

-   Stressful DO: 2-5 mg/L

-   Hypoxic DO: \<2 mg/L

#### Surface Water Temperature Module

Input: Daily lake surface water temperature (ºC)

-   Processing: Daily values aggregated to monthly means by water year (October - September)

-   Function: temperature_function

Output: Mean Lake Temperature Classifications:

-   Cool: Below Average \<13ºC

-   Mild: Average 13ºC - 22ºC

-   Warm: Above Average 22ºC - 30ºC

-   Hot: Boiling \>30ºC

### Lake Water Quality Module Parameters

Inputs: Total productivity (mg/L ), mean dissolved oxygen (mg/L ), and mean surface water temperature (ºC)

-   Expected inputs: 365 daily observations per water year (October - September)

-   Functions used: temperature_function, nutrients_function, and dissoxygen_function

-   Processing: Monthly classifications from each module combined to assign overall water quality risk

Outputs: water_quality_model - Monthly water quality risk classifications

-   Low Risk - Oligotrophic, Excellent DO, Cool temperature

-   Moderate Risk - Mesotrophic, Moderate DO, Mild temperature

-   High Risk - Eutrophic, Stressful DO, Warm temperature

-   Extreme Risk - Hypereutrophic, Hypoxic DO, Hot temperature

## Flow Chart / Diagram

```{r echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
library(here)
library(htmltools)
library(base64enc)

pdf_base64 <- dataURI(
  file = here::here("figures", "flowchart.pdf"),
  mime = "application/pdf"
)

pdf_link <- tags$a(
  href = pdf_base64,
  download = "flowchart.pdf",
  "Download Full Flow Chart PDF"
)

flowchart_img <- tags$img(
  src = knitr::image_uri(here::here("figures", "flowchart.jpg")),
  style = "width:100%; border:1px solid #ddd;"
)

tagList(
  pdf_link,
  tags$br(),
  tags$br(),
  flowchart_img
)
```

<<<<<<< HEAD
# Water Quality Model
=======
## Final Lake Water Quality Model
>>>>>>> ba36579152678c2578c385ef0278e308149615a6

```{r}
# Script for Final Water Quality Model

# loading libraries
library(tidyverse)
library(here)

# load in the model functions
source(here("scripts", "nutrient_module_script.R"))
source(here("scripts", "dissoxygen_module_script.R"))
source(here("scripts", "temperature_module_script.R"))

# load in the data
nutrientdata <- read.csv(here("data", "nutrient_data.csv"))
dissoxygendata <- read.csv(here("data", "dissoxygen_data.csv"))
temperaturedata <- read.csv(here("data", "temperature_data.csv"))

# run each model
nutrient_results <- nutrient_function(nutrientdata)
dissoxygen_results <- dissoxygen_function(dissoxygendata)
temperature_results <- temperature_function(temperaturedata)

# combine model outputs by month
water_quality_data <- nutrient_results |>
  select(month, nutrient_class = classification) |>
  left_join(
    dissoxygen_results |>
      select(month, dissoxygen_class = classification),
    by = "month"
  ) |>
  left_join(
    temperature_results |>
      select(month, temperature_class = classification),
    by = "month"
  )

# final water quality classification function
water_quality_model <- function(nutrient_class,
                                dissoxygen_class,
                                temperature_class) {
  
  if (nutrient_class == "Hypereutrophic" ||
      dissoxygen_class == "Hypoxic" ||
      temperature_class == "Hot - Boiling") {
    
    final_class <- "Extreme Risk"
    
  } else if (nutrient_class == "Eutrophic" ||
             dissoxygen_class == "Stressful" ||
             temperature_class == "Warm - Above Average") {
    
    final_class <- "High Risk"
    
  } else if (nutrient_class == "Mesotrophic" ||
             dissoxygen_class == "Moderate" ||
             temperature_class == "Mild - Average") {
    
    final_class <- "Moderate Risk"
    
  } else {
    
    final_class <- "Low Risk"
  }
  
  return(final_class)
}

# apply final model
water_quality_results <- water_quality_data |>
  mutate(
    final_water_quality = mapply(
      water_quality_model,
      nutrient_class,
      dissoxygen_class,
      temperature_class
    )
  )

# show final results
print(water_quality_results)
```

## Lake Water Quality Visualization Module

```{r}

# Water Quality Visualization Script

library(tidyverse)
library(here)
# -------------------------------------------------------------------
# Plot 1: Monthly Final Water Quality Classification
#
# This figure displays the overall monthly lake water quality risk
# classification produced by the integrated water quality model.
# -------------------------------------------------------------------

# Set factor order for water quality classifications
water_quality_results$final_water_quality <- factor(
  water_quality_results$final_water_quality,
  levels = c(
    "Low Risk",
    "Moderate Risk",
    "High Risk",
    "Extreme Risk"
  )
)


# Create visualization
wqm_classification_bargraph <- ggplot(
  water_quality_results,
  aes(
    x = month,
    y = final_water_quality,
    fill = final_water_quality
  )
) +
  geom_col(color = "black", width = 0.7) +
  labs(
    title = "Monthly Final Lake Water Quality Classification By Water Year",
    x = "Month",
    y = "Water Quality Risk Classification",
    fill = "Risk Level"
  ) +
  scale_fill_manual(values = c(
    "Low Risk" = "#66c2a5",
    "Moderate Risk" = "#ffd166",
    "High Risk" = "#f4a261",
    "Extreme Risk" = "#d62828"
  )) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    axis.title = element_text(face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.title = element_text(face = "bold")
  )


wqm_classification_bargraph
```

```{r}

# -------------------------------------------------------------------
# Plot 2: Monthly Environmental Conditions by Module
#
# This figure displays the monthly mean phosphorus, nitrogen,
# dissolved oxygen, and temperature values used in the lake
# water quality model. The plot highlights seasonal environmental
# patterns across the water year.
# -------------------------------------------------------------------

# Combine monthly model outputs
visual_data <- nutrient_results |>
  select(month,
         mean_phosphorus,
         mean_nitrogen) |>
  left_join(
    dissoxygen_results |>
      select(month, mean_dissoxygen),
    by = "month"
  ) |>
  left_join(
    temperature_results |>
      select(month, mean_temp_c),
    by = "month"
  )

# Convert to long format for plotting
visual_data_long <- visual_data |>
  pivot_longer(
    cols = c(mean_phosphorus,
             mean_nitrogen,
             mean_dissoxygen,
             mean_temp_c),
    names_to = "variable",
    values_to = "value"
  )

# Create visualization
monthly_conditions_linegraph <- ggplot(visual_data_long,
       aes(x = month,
           y = value,
           color = variable,
           group = variable)) +
  geom_line(size = 1.2) +
  geom_point(size = 3) +
  labs(
    title = "Monthly Lake Environmental Conditions By Water Year",
    x = "Month",
    y = "Mean Monthly Value",
    color = "Environmental Variable"
  ) +
  scale_color_manual(values = c(
    "mean_phosphorus" = "darkorange",
    "mean_nitrogen" = "forestgreen",
    "mean_dissoxygen" = "steelblue",
    "mean_temp_c" = "firebrick"
  ),
  labels = c(
      "mean_phosphorus"  = "Phosphorus (mg/L)",
      "mean_nitrogen"    = "Nitrogen (mg/L)",
      "mean_dissoxygen"  = "Dissolved Oxygen (mg/L)",
      "mean_temp_c"      = "Surface Water Temperature (°C)")) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

monthly_conditions_linegraph
```
