---
title: "Lake Eutrophication and Water Quality Classification Model"
author: "Lauren Chetley & Emma Rodriguez"
format: html
editor: visual
---

## Contract

### Parameters

#### Temperature

Temperature Module: surface water temperature (swt), seasonal time (temp.time)

Temperature Classification:

-   Cool

-   Mild

-   Average

-   Warm

-   Hot

#### Nutrient

Nutrient Module: nitrogen (n.input), phosphorus (p.input), seasonal time (nutrient.time)

Productivity Risk Classification

-   Oligotrophic

-   Mesotrophic

-   Eutrophic

-   Hypereutrophic

#### Dissolved Oxygen

Dissolved Oxygen Module: dissolved oxygen (diss.oxy), seasonal time (diss.oxy.time)

-   Excellent

-   Moderate

-   Stressful

-   Hypoxic

### Inputs & Outputs

Inputs: surface water temperature, temperature seasonal time, nitrogen, phosphorus, nutrient seasonal time, dissolved oxygen, dissolved oxygen seasonal time

Outputs: Water Quality Model

-   Low Risk - Oligotrophic, excellent, cool

-   Moderate Risk - Mesotrophic, moderate, mild

-   High Risk - Eutrophic, stressful, warm

-   Extreme Risk - Hypereutrophic, hypoxic, hot

### Flow Chart / Diagram

```{r}


```

## Final Water Quality Model

```{r}
# Script for Final Water Quality Model

# loading libraries
library(tidyverse)
library(here)

# load in the model functions
source(here("scripts", "nutrient_module_script.R"))
source(here("scripts", "dissoxygen_module_script.R"))
source(here("scripts", "temperature_module_script.R"))

# load in the data
nutrientdata <- read.csv(here("data", "nutrient_data.csv"))
dissoxygendata <- read.csv(here("data", "dissoxygen_data.csv"))
temperaturedata <- read.csv(here("data", "temperature_data.csv"))

# run each model
nutrient_results <- nutrient_function(nutrientdata)
dissoxygen_results <- dissoxygen_function(dissoxygendata)
temperature_results <- temperature_function(temperaturedata)

# combine model outputs by month
water_quality_data <- nutrient_results |>
  select(month, nutrient_class = classification) |>
  left_join(
    dissoxygen_results |>
      select(month, dissoxygen_class = classification),
    by = "month"
  ) |>
  left_join(
    temperature_results |>
      select(month, temperature_class = classification),
    by = "month"
  )

# final water quality classification function
water_quality_model <- function(nutrient_class,
                                dissoxygen_class,
                                temperature_class) {
  
  if (nutrient_class == "Hypereutrophic" ||
      dissoxygen_class == "Hypoxic" ||
      temperature_class == "Hot - Boiling") {
    
    final_class <- "Extreme Risk"
    
  } else if (nutrient_class == "Eutrophic" ||
             dissoxygen_class == "Stressful" ||
             temperature_class == "Warm - Above Average") {
    
    final_class <- "High Risk"
    
  } else if (nutrient_class == "Mesotrophic" ||
             dissoxygen_class == "Moderate" ||
             temperature_class == "Mild - Average") {
    
    final_class <- "Moderate Risk"
    
  } else {
    
    final_class <- "Low Risk"
  }
  
  return(final_class)
}

# apply final model
water_quality_results <- water_quality_data |>
  mutate(
    final_water_quality = mapply(
      water_quality_model,
      nutrient_class,
      dissoxygen_class,
      temperature_class
    )
  )

# show final results
print(water_quality_results)
```

# Water Quality Visualization Module

```{r}

# Water Quality Visualization Script

library(tidyverse)
library(here)

# -------------------------------------------------------------------
# Plot 1: Monthly Final Water Quality Classification
#
# This figure displays the overall monthly lake water quality risk
# classification produced by the integrated water quality model.
# -------------------------------------------------------------------

wqm_classification_bargraph <- ggplot(water_quality_results,
       aes(x = month,
           y = final_water_quality,
           fill = final_water_quality)) +
  geom_col(color = "black", width = 0.7) +
  labs(
    title = "Monthly Final Water Quality Classification",
    x = "Month",
    y = "Water Quality Risk Classification",
    fill = "Risk Level"
  ) +
  scale_fill_manual(values = c(
    "Low Risk" = "#66c2a5",
    "Moderate Risk" = "#ffd166",
    "High Risk" = "#f4a261",
    "Extreme Risk" = "#d62828"
  )) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    axis.title = element_text(face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.title = element_text(face = "bold")
  )

wqm_classification_bargraph

```

```{r}

# -------------------------------------------------------------------
# Plot 2: Monthly Environmental Conditions by Module
#
# This figure displays the monthly mean phosphorus, nitrogen,
# dissolved oxygen, and temperature values used in the lake
# water quality model. The plot highlights seasonal environmental
# patterns across the water year.
# -------------------------------------------------------------------

# Combine monthly model outputs
visual_data <- nutrient_results |>
  select(month,
         mean_phosphorus,
         mean_nitrogen) |>
  left_join(
    dissoxygen_results |>
      select(month, mean_dissoxygen),
    by = "month"
  ) |>
  left_join(
    temperature_results |>
      select(month, mean_temp_c),
    by = "month"
  )

# Convert to long format for plotting
visual_data_long <- visual_data |>
  pivot_longer(
    cols = c(mean_phosphorus,
             mean_nitrogen,
             mean_dissoxygen,
             mean_temp_c),
    names_to = "variable",
    values_to = "value"
  )

# Create visualization
monthly_conditions_linegraph <- ggplot(visual_data_long,
       aes(x = month,
           y = value,
           color = variable,
           group = variable)) +
  geom_line(size = 1.2) +
  geom_point(size = 3) +
  labs(
    title = "Monthly Environmental Conditions",
    x = "Month",
    y = "Mean Monthly Value",
    color = "Environmental Variable"
  ) +
  scale_color_manual(values = c(
    "mean_phosphorus" = "darkorange",
    "mean_nitrogen" = "forestgreen",
    "mean_dissoxygen" = "steelblue",
    "mean_temp_c" = "firebrick"
  )) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

monthly_conditions_linegraph
```
