---
title: "Lake Eutrophication and Water Quality Classification Model"
author: "Lauren Chetley & Emma Rodriguez"
format: html
editor: visual
---

## Contract

### Parameters

#### Temperature

Temperature Module: surface water temperature (swt), seasonal time (temp.time)

Temperature Classification:

-   Cool

-   Mild

-   Average

-   Warm

-   Hot

#### Nutrient

Nutrient Module: nitrogen (n.input), phosphorus (p.input), seasonal time (nutrient.time)

Productivity Risk Classification

-   Oligotrophic

-   Mesotrophic

-   Eutrophic

-   Hypereutrophic

#### Dissolved Oxygen

Dissolved Oxygen Module: dissolved oxygen (diss.oxy), seasonal time (diss.oxy.time)

-   Excellent

-   Moderate

-   Stressful

-   Hypoxic

### Inputs & Outputs

Inputs: surface water temperature, temperature seasonal time, nitrogen, phosphorus, nutrient seasonal time, dissolved oxygen, dissolved oxygen seasonal time

Outputs: Water Quality Model

-   Low Risk - Oligotrophic, excellent, cool

-   Moderate Risk - Mesotrophic, moderate, mild

-   High Risk - Eutrophic, stressful, warm

-   Extreme Risk - Hypereutrophic, hypoxic, hot

### Flow Chart / Diagram

```{r}


```

## Final Water Quality Model

```{r}
# Script for Final Water Quality Model

# loading libraries
library(tidyverse)
library(here)

# load in the model functions
source(here("scripts", "nutrient_model.R"))
source(here("scripts", "dissoxygen_model.R"))
source(here("scripts", "temperature_model.R"))

# load in the data
nutrientdata <- read.csv(here("data", "nutrient_data.csv"))
dissoxygendata <- read.csv(here("data", "dissoxygen_data.csv"))
temperaturedata <- read.csv(here("data", "temperature_data.csv"))

# run each model
nutrient_results <- nutrient_model(nutrientdata)
dissoxygen_results <- dissoxygen_model(dissoxygendata)
temperature_results <- temperature_model(temperaturedata)

# combine model outputs by month
water_quality_data <- nutrient_results |>
  select(month, nutrient_class = classification) |>
  left_join(
    dissoxygen_results |> select(month, dissoxygen_class = classification),
    by = "month"
  ) |>
  left_join(
    temperature_results |> select(month, temperature_class = classification),
    by = "month"
  )

# final water quality classification
water_quality_model <- function(nutrient_class, dissoxygen_class, temperature_class) {
  
  if (nutrient_class == "Oligotrophic" &&
      dissoxygen_class == "Excellent" &&
      temperature_class == "Cool") {
    final_class <- "Low Risk"
    
  } else if (nutrient_class == "Mesotrophic" &&
             dissoxygen_class == "Moderate" &&
             temperature_class == "Mild") {
    final_class <- "Moderate Risk"
    
  } else if (nutrient_class == "Eutrophic" &&
             dissoxygen_class == "Stressful" &&
             temperature_class == "Warm") {
    final_class <- "High Risk"
    
  } else if (nutrient_class == "Hypereutrophic" ||
             dissoxygen_class == "Hypoxic" ||
             temperature_class == "Hot") {
    final_class <- "Extreme Risk"
    
  } else {
    final_class <- "Mixed Risk"
  }
  
  return(final_class)
}

# apply final function to each month
water_quality_results <- water_quality_data |>
  mutate(
    final_water_quality = mapply(
      water_quality_model,
      nutrient_class,
      dissoxygen_class,
      temperature_class
    )
  )

# show final results
print(water_quality_results)
```

# Water Quality Visualization Module

```{r}

```
